---
title: "A study of on-ramp metering in Perth"
format:
  html:
    css: styles.css
editor: visual
---

| Name              | Student ID |
|-------------------|------------|
| Christian Fazzari | 22220880   |
| Annas Atif        | 22224125   |
| Madhan Jeelaga    | 22815626   |

: *Project Authors*

## Introduction and Background (10 marks)

-   Problem context

-   Objectives

-   Significance

-   Brief literature review (3-5 relevant sources)

Background \<- reid highway is arguably one of the most important links in with the mitchell freeway (only highway-freeway interchange, all others are residential avenues, important for transit of goods + people)

Perth's freeway network has undergone major changes in recent years. One of the more controversial upgrades has been the Smart Freeways project; a modification which aims to improve the travel time of commuters from Perth's outer suburbs heading towards the city. This system has been implemented on both the Kwinana Freeway Northbound and the Mitchell Freeway Southbound, controlling the flow of traffic predominantly through digital speed-limit gantries and on-ramp metering. On the Mitchell South, this system saves 7-minutes when travelling from Hester Avenue to Vincent Street in the morning peak (Main Roads, 2025). However, residents closer to the city are generally more impacted by ramp-metering, and, have noticed their morning commute more than doubling in length (ABC, 2025).

In order to weigh the validitity of these claims, this study will analyse the time spent queuing the enter the freeway and compare this with previous research of the time saved on the mainline. We will perform parameter optimisation to adjust and maximise release timing while maintaining stability of the system.

A queuing system with $c$ lanes is stable (i.e. has a controllable load) if its utilisation rate $\rho = \frac{\lambda}{c \times \mu}$ is less than $1$, where $\lambda$ and $\mu$ refer to the mean arrival and departure rates respectively (Righini, 2020). In terms of ramp-metering, this utilisation rate is typically maintained in a range that is safely below capacity as a means of risk aversion, or avoiding a total breakdown of the queue system (Levinson, 2021).

**The primary aim of this study is the determine whether the reported 7-minute travel time savings on the Kwinana Freeway Northbound and Mitchell Freeway Southbound are negated by the increased on-ramp queuing times for local residents. This will be evaluated by developing a queuing simulation based on historical traffic volume data.**

<https://pages.cs.wisc.edu/~dsmyers/cs547/simulation_handout.pdf>

\*\* this is unfinished but just trying to get ideas

```{r "Importing and filtering"}
#| include: false

# Keep these commented as the full files were too large to upload to Github
# Importing datasets
# raw23 <- read.csv("2023-03.csv")
# raw25 <- read.csv("2025-03.csv")

# Filtering for on-ramp segment
# ramp25 <- raw25[raw25$MLinkID == 5725, ]
# ramp23 <- raw23[raw23$MLinkID == 5725, ]

# Filtering for the freeway segment directly following the on-ramp
# fwy25 <- raw25[raw25$MLinkID == 183, ]
# fwy23 <- raw23[raw23$MLinkID == 183, ]

# Simplified import commands
ramp25 <- read.csv("ramp25.csv")
ramp23 <- read.csv("ramp23.csv")

fwy25 <- read.csv("fwy25.csv")
fwy23 <- read.csv("fwy23.csv")
```

```{r "Date and time modifications"}
#| include: false

library(lubridate)

# Converting datetime strings to date objects
ramp23$Date.Time.AWST <- ymd_hm(ramp23$Date.Time.AWST)

# Adding day of the week column (1=sunday, 7=saturday)
ramp23$Weekday <- wday(ramp23$Date.Time.AWST)

# Adding hour of the day column
ramp23$Hour <- hour(ramp23$Date.Time.AWST)

# Adding 15-min segment identifier of column
ramp23$Period <- format(ramp23$Date.Time.AWST, "%H:%M")

# Removing weekends
ramp23.wk <- ramp23[ramp23$Weekday %in% c(2, 3, 4, 5, 6), ]

# Filtering for only the morning peak (6am-10am)
ramp23.pk.wk <- ramp23.wk[6 <= ramp23.wk$Hour & ramp23.wk$Hour < 10, ]

# Removing public holidays (March 6 is the onyl date - Labour Day)
ramp23.pk.wk <- ramp23.pk.wk[date(ramp23.pk.wk$Date.Time.AWST) != "2023-03-06", ]
```

## Methodology (25 marks)

-   Assumptions and model formulation

-   Description of data (if applicable)

-   Justification of modelling choices

Main Roads have a large online repository of traffic volume measurements from 2013 to the present day. Each entry is a 15-minute interval which can be accessed for any road segment using its unique MLinkID. We can filter for the on-ramp of interest using its ID 5725:

```{r "Filtered datset"}
head(ramp23[, c(-6, -7, -8, -9, -10)])
```

> Here, 'Volume' refers to the count of vehicles, and 'TravelTimeMinutes' the average time to travel that particular road segment.

TALK ABOUT MEMORYLESS PROPERTY AS A FALLBACK/ASSUMPTION

In order to construct a queuing model, we can use this dataset to simulate the incoming volume of traffic during the morning peak. Typically, this model would be formed on the Poisson distribution because of its memoryless property and ease of implementation.

> It is important to discuss the memoryless property of the poisson process as this implies that the probability of future events is completely **independent** of the past as the process is unable to 'mathematically determine' the time that has passed since the last event. This makes it an ideal process to use as we can model extreme conditions such as traffic accidents whilst ensuring that they remain completely independent of the other events.

However, a Poisson distribution assumes an equal mean and variance, so it's important that verify this assumption. To visualise, let's filter the above for weekdays, the peak hours, and without public holidays (eg 06/03/23 : Labour Day), yielding the following:

```{r "Aggregating average daily volumes"}
#| include: false

smeans <- aggregate(Volume ~ Period, 
                           data = ramp23.pk.wk, 
                           FUN = mean)

# getting max volume
smax <- max(smeans$Volume)
maxindex <- which(smeans$Volume == smax)

# getting mean volume
smean <- mean(smeans$Volume)

# getting min volume
smin <- min(smeans$Volume)
minindex <- which(smeans$Volume == smin)
```

```{r "Volume barplot"}
#| echo: false

bp <- barplot(smeans$Volume, names.arg = smeans$Period, col="#abbf9d", xlab = "Time of Morning", ylab = "Mean Volume of Period", main="Volume of Merging Traffic, March 2023")

abline(v = bp[1], col = "#252e1f", lwd = 3, lty = 2)
text(x = bp[1], y = smax, labels = round(smeans$Volume[1], 2), pos = 3, xpd = TRUE)

abline(v = bp[5], col = "#252e1f", lwd = 3, lty = 2)
text(x = bp[5], y = smax, labels = round(smeans$Volume[5], 2), pos = 3, xpd = TRUE)

abline(v = bp[9], col = "#252e1f", lwd = 3, lty = 2)
text(x = bp[9], y = smax, labels = round(smeans$Volume[9], 2), pos = 3, xpd = TRUE)
```

The volume of traffic entering the freeway varies heavily over the course of the morning. It would be unreasonable to fit a single model to this level of variation, so instead consider a piecewise approach. Models that account for the minimum, maximum and mean volume each morning will be constructed, correlating to the times 6am, 7am and approximately 8am.

We'll focus on the 6:00-6:15am interval for now:

```{r "Parameters for 6am model"}
#| echo: false
smu6  <- mean(ramp23.pk.wk[ramp23.pk.wk$Period == "06:00", 4])
svar6 <- var(ramp23.pk.wk[ramp23.pk.wk$Period == "06:00", 4])

data.frame (
  Mean = smu6,
  Variance = svar6,
  row.names = c("6am (min)")
)
```

This is an overdispersed dataset where the variance of volume is considerably higher than its mean. This is true throughout the entire dataset, meaning a Poisson distribution would be unsuitable as we do not satisfy the assumption of equidispersion. In order to provide a better fit, we can consider a Negative Binomial where both a mean and a size parameter (for dispersion) can be selected. (NEED TO TALK ABOUT HOW MEMORYLESS IS STILL INVALID AND ALSO ASSUMPTION OF INDEPENDENCE REMAINS)

> It is important to note that, whilst we eliminate the memoryless aspect of the Poisson Process, we still retain independence with the Negative Binomial distribution. In fact, it makes a better proposition compared to the Poisson Process, as we are able to handle overdispersion in the data set which allows for more accurate modelling of accidents, and in high congestion traffic zones which makes it ideal for this studies case.

We can calculate this size parameter $r$ as follows:

$$r = \frac{\mu^2}{\sigma^2 - \mu}$$

```{r "Calculating 6am disperion factor"}
#| include: false
ssize6 <- (smu6^2) / (svar6 - smu6)
```

To see the improvement in fit, let's simulate a Poisson and a Negative Binomial with these parameters and compare with the original dataset.

```{r "Simulating goodness of fit improvement"}
sim1 <- rnbinom(10000, size = ssize6, mu = smu6)
sim2 <- rpois(10000, lambda = smu6)
data.frame (
  Mean = c(smu6, mean(sim1), mean(sim2)),
  Variance = c(svar6, var(sim1), mean(sim2)),
  row.names = c("Dataset", "NBinom Sim", "Poisson Sim  ")
)
```

The Negative Binomial is a much better fit to the original dataset. However, a new concern arises: for the purpose of simulation, we wish to simulate vehicle-counts on a per-minute basis. This will help to increase the granularity of our queuing time results. One possible solution (*Method 1*) is to simply divide the volume data by 15mins and fit a new model. If we consider this case, our mean will scale by a factor of $\frac{1}{15}$, and the variance by a factor of $\frac{1}{15^2}$:

```{r "Method 1 of reducing to Vehicles/Min"}
#| echo: false
data.frame(
  Mean = smu6 / 15,
  Variance = svar6 / 15^2,
  Size = ((smu6/15)^2) / ((svar6/(15^2)) - smu6/15),
  row.names = c("Vehicles/min")
  )
```

Since this is now underdispersed, a Negative Binomial will no longer fit (hence the negative size). It remains unbiased, but erases so much of the original variance that it is completely irrelevant. An alternative solution (*Method 2*) that maintains our original variation is to use the divisible property of the Negative Binomial. We first form the model, then scale its parameters to maintain a dispersion that remains in proportion to the new, lower mean.

> This occurs because the Negative Binomial is formed as a Compound Poisson (Siegrist, 2022), so if $X=X_1 + X_2 +...+X_n \sim NBinom(k, p)$ then we can model each $X_i$ as $NBinom(\frac{k}{n}, p)$ with the same probability $p$. Extrapolating this $(_{successes}, \space _{probability})$ form to the $(_{mean}, \space _{size})$ form yields the parameters $(\frac{\mu}{n}, \space \frac{r}{n})$.

```{r Compare methods 1 and 2 of reducing to Vehicles/Min}
#| echo: false

data.frame(
  Mean = c(smu6, smu6/15, smu6/15),
  Size = c(ssize6, ((smu6/15)^2) / ((svar6/(15^2)) - smu6/15), ssize6/15),
  row.names = c("Vehicles/15min", "Vehicles/min (Method 1)", "Vehicles/min (Method 2)")
)
```

We can extend this idea from the 6am model to all three models.

------------------------------------------------------------------------

::: caption
Disaggregated Model Parameters (vehicles per min)
:::

```{r "Updating 6am and calculating 7am and 8am parameters to Vehicles/Min"}
#| include: false

# 6am models
# leave as the same

# 7am models
smu7  <- mean(ramp23.pk.wk[ramp23.pk.wk$Period == "07:00", 4])
svar7 <- var(ramp23.pk.wk[ramp23.pk.wk$Period == "07:00", 4])
ssize7 <- ((smu7^2) / (svar7 - smu7))

# 8am models
smu8  <- mean(ramp23.pk.wk[ramp23.pk.wk$Period == "08:00", 4])
svar8 <- var(ramp23.pk.wk[ramp23.pk.wk$Period == "08:00", 4])
ssize8 <- ((smu8^2) / (svar8 - smu8))

# update to vehicles / 15 minute
# we no longer use variance! DO NOT TOUCHHHH VARIANCE
smu6 <- smu6 / 15
ssize6 <- ssize6 / 15

smu7 <- smu7 / 15
ssize7 <- ssize7 / 15

smu8 <- smu8 / 15
ssize8 <- ssize8 / 15
```

```{r "Showing parameters for all models"}
#| echo: false

data.frame(
  Mean = c(smu6, smu7, smu8),
  Size = c(ssize6, ssize7, ssize8),
  row.names = c("6am (min)", "7am (max)", "8am (mean)")
)
```

## Model Implementation and Results (45 marks)

Primary points:

-   Implementation (code integrated with explanation)

-   Statistical summaries and visualisations

-   Interpretation in context

Secondary points:

-   Extend to multiple servers (M/M/c) and compare

-   Non-Exponential Service Times (M/G/1)

-   Compare simulation results with analytical formulas (where available)

-   Investigate waiting time distributions

-   Time-varying arrivals (peak vs off-peak) using piecewise arrival rates

-   Conduct sensitivity analysis on arrival and service rates

For the simulation, we can develop a series of queuing models based on the parameters described above. We have developed a reusable function which can represent different arrival and release rates, applied for an G/M/c queue with Negative Binomial arrivals.

```{r "Primary Simulation"}
onRampSimulation <- function(steps, alambda, asize, rmu, c) {
  set.seed(1234)
  
  queue <- c() # storage for vehicles currently in queue
  incoming <- numeric(steps) # track count of vehicles arrived
  outgoing <- numeric(steps) # track count of vehicles released
  
  wlength <- numeric(steps) # track length of queue in each timestep
  wtimesavg <- numeric(steps) # track AVERAGE waiting times of vehicles released in a timestep
  wtimesall <- c() # track INDIVIDUAL waiting times of vehicles released in a timestep

  for (i in 1:steps) {
    # calculate arrivals
    arate <- rnbinom(1, size=asize, mu=alambda)
    
    queue <- c(queue, rep(i, arate))
    incoming[i] <- arate
    
    # calculate departures
    rrate <- sum(rpois(c, lambda=rmu))
    released <- min(length(queue), rrate)
    
    outgoing[i] <- released
    
    # release cars and track wait times
    if (released > 0) {
      rtimes <- i - queue[1:released]
      wtimesavg[i] <- mean(rtimes)
      wtimesall <- c(wtimesall, rtimes)
      queue <- queue[-(1:released)]
    } else {
      wtimesavg[i] <- 0 
    }
    
    wlength[i] <- length(queue)
  }
  
  return(list(
    incoming = incoming,
    outgoing = outgoing,
    wtimesavg = wtimesavg,
    wtimesall = wtimesall,
    wlength = wlength)
    )
}
```

```{r "Multi-Sim Function"}
#| include: false

# easy method to run mutliple simulations
runMultiSims <- function(steps, alambdas, asizes, rmus, c){
  if (length(rmus) != length(alambdas) || length(alambdas) != length(asizes))
    stop("Must have an equal count for each parameter.")
  
  sims <- list()

  for (i in seq_along(rmus)) {
    sims[[i]] <- onRampSimulation(
      steps = steps, 
      alambda = alambdas[i],
      asize = asizes[i],
      rmu = rmus[i], 
      c = c
    )
  }
  
  maxlength <- 0
  for (sim in sims) {
    maxlength <- max(sim$wlength, maxlength)
  }
  
  data <- list(
    sims = sims,
    maxlength = maxlength
  )
  
  return(data)
}
```

### Part 1 - Sensitivity Analysis

For each model we have developed, consider the critical point where the utilisation parameter $\rho$ is exactly one. Rearranging in terms of the release rate, we find $\mu = \frac{\lambda}{1 \times c}$ whichs allows us to determine the largest timing possible for each system to remain stable. As 2 lanes are present at the Reid Highway on-ramp, we set $c=2$ and form a G/M/2 queue.

```{r "Calculating departure rates"}
#| echo: false

c <- 2

rmu6 <- smu6/c
rmu7 <- smu7/c
rmu8 <- smu8/c

data.frame(
  Best = c(smu6, rmu6),
  Worst = c(smu7, rmu7),
  Average = c(smu8, rmu8),
  row.names = c("Arrival Rate", "Departure Rate")
)
```

We can start by running the above simulation for the minimum (best, 6am), maximum (worst, 7am) and mean (average, 8am) models:

::: caption
Best, Worst and Average Case Simulations (at critical points)
:::

```{r "Visualising Stability"}
#| echo: false
alambdas <- c(smu6, smu7, smu8)
asizes <- c(ssize6, ssize7, ssize8)
rmus <- c(rmu6, rmu7, rmu8)

sims <- runMultiSims(10000, alambdas, asizes, rmus, 2)
maxqueue <- sims$maxlength
sims <- sims$sims # can just rassign

par(mfrow=c(1, length(rmus)), mar=c(4, 4, 2, 1))

for (i in seq_along(rmus)) {
  plot(
    sims[[i]]$wlength,
    type = "l",
    xlab = "Timestep",
    ylab = "Count of Vehicles",
    ylim = c(0, maxqueue),
    cex.axis = 0.8,
    col = "#b5b5b5",
    main = paste("Lambda =", round(alambdas[i], 2), ", Mu =", round(rmus[i], 2))
  )
}

paste("Maximum length:", maxqueue)

```

While in all cases there is theoretical stability (i.e. the models will all return to 0), we can see that a utilisation rate of exactly $1$ is completely unreasonable in a real-world scenario. The 7am model particularly balloons to a queue of just under 1200 vehicles long, with the others reaching more than 400 vehicles each. In reality, a utilisation rate closer to 90% is generally maintained to balance throughput with wait times.

Focusing on the worst-case scenario (7am), we can reaffirm the visualisions through a plot of the ratio of arrival-to-depature counts for a series of release rates $\mu$. This ratio can be $1$ at minimum as we can only have as many vehicles departing as there were vehicles that have arrived in the first place. However, when the system is past stability this ratio is unbounded.

```{r "7am Analysis"}
#| echo: false

steps <- 1000
rmus <- seq(2, 15)
count <- length(rmus)
c <- 2

sims <- runMultiSims(steps, rep(smu7, count), rep(ssize7, count), rmus, c)
# inoutrate <- sapply(sims$sims, function(x) mean(x$incoming/x$outgoing))
inoutrate <- sapply(sims$sims, function(x) sum(x$incoming)/sum(x$outgoing))
meanlength <- sapply(sims$sims, function(x) mean(x$wlength)/steps) # /steps to normalise

par(mfrow=c(1, 2))
plot(rmus, inoutrate, main="Incoming/Outgoing Ratio", ylab="Rate Parameter", xlab="Mu", pch=16, col="#b5b5b5")
abline(v=8, col="red", lwd = 2)
plot(rmus, meanlength, col="#b5b5b5", main="Normalised Queue Length", ylab="Length of Queue / Timesteps", xlab="Mu")
abline(v=8, col="red", lwd = 2)

paste("Recall that the incoming rate is", round(smu7, 2), "with a critical Mu of", round(rmu7, 2))
```

For $\mu < 8$ there is an unbounded exponential growth in the ratio of incoming-to-outgoing vehicles as was expected. For $\mu \geq 8$, this ratio is clamped at one (ignoring random variation). A larger ratio means there are consistently more vehicles arriving than the system can output, and so if this is above one total breakdown will occur.

This suggests that the queuing model is highly sensitive to changes in its departure rate past the critical utilisation rate $\rho \geq 1$. To investigate further, we can shift focus to the normalised queue length plot on the right. Here, notice the linear shape past the same critical point. We can define this region by the line:

```{r "Above 1 Sensitivity"}
slope <- lm(meanlength[rmus < 8] ~ rmus[rmus < 8], )
summary(slope)
```

We find a slope of $-1.002 \approx -1$, suggesting that for every unit increase in $\mu$ getting closer to the critical point, we lose a unit of length in the queue. So interestingly, while the ratio parameter seems to grow exponentially, the relationship between the resulting length of the queue for a given release rate $\mu$ is linear.

Let's now shift focus to the sensitivity of the system while $\rho < 1$.

```{r "7am Analysis Continued"}
#| echo: false

steps <- 5000
rmus <- c(7.5, round(rmu7, 4), seq(8, 9, 0.25))
count <- length(rmus)
c <- 2

# can delete shite above
cols <- hcl.colors(length(rmus), "Viridis")

sims <- runMultiSims(steps, rep(smu7, count), rep(ssize7, count), rmus, c)$sims

plot(
    cumsum(sims[[1]]$incoming) / cumsum(sims[[1]]$outgoing),
    type = "l",
    xlab = "Timestep",
    ylab = "Ratio Value",
    cex.axis = 0.8,
    col = cols[1],
    main = "Cumulative Incoming / Cumulative Outgoing"
  )

# need to replot 1 so it shows ontop
for (i in seq_along(rmus)) {
  lines(
    cumsum(sims[[i]]$incoming) / cumsum(sims[[i]]$outgoing),
    col=cols[i]
    )
}

legend("topright", legend = rmus, col = cols, lty = 1, lwd = 2, title = "Mu Values", bty = "n", cex = 0.8)
```

As before, we need this to trend near $1.00$ for long-term timesteps, otherwise there are consistently more vehicles arriving than leaving. This seems to occur for all but the smallest two $\mu$ values which are clearly disjoint from the rest. We can therefore say that for $\mu > 7.76$, the lines appear to converge and return back to $1.00$.

```{r "Below 1 Sensitivity"}
#| echo: false

# use sims from above

rmus <- rmus[c(-1, -2)]
sims <- sims[c(-1, -2)]

plotlist <- lapply(sims, function(x) x$wtimesavg)
names(plotlist) <- rmus

boxplot(plotlist, 
        horizontal=TRUE, 
        las=1,
        col=hcl.colors(length(rmus), "Viridis"),
        xlab="Wait Time (min)", 
        ylab="Release Rate (Mu)",
        main="Wait Time Variability across Mu Values",
        outline=TRUE,
        cex=0.8
        )
```

> This is a plot of average wait-times for each timestep, allowing us to achieve sub-1-minute resolution. This slighly reduces the maximum wait-times seen but greatly aides the visualisation.

There is a clear exponential growth for the boxplots as $\mu \rightarrow 8$, or in other terms $\rho \rightarrow 1$:

-   the median wait time for vehicles increases;

-   it's IQR widens; so we have more uncertainty if quoting a given mean

-   the whiskers grow far longer; so outlier vehicles will notice a disproportionally extreme increase in wait time

Choosing to ignore the outliers, a reasonable choice that fits everyday, predicable traffic would be $\mu=8.25$.

```{r}
summary(plotlist$"8.25")
```

Here, 75% of drivers are waiting under $2.2$ minutes with a median of $1$ minute. However, notice the maximum wait-time of $16$ minutes. This system would benefit from live feedback, as it can maintain a smaller $\mu$ when under regular operation, but if such a spike occurs it can adjust to prevent the queue from ballooning as is occuring here.

Instead considering the outliers, it takes until $\mu = 8.75$ for there to be no appreciable decrease in wait time at the on-ramp as $\mu$ decreases further, corresponding to a utilisation of $\frac{15.5}{2 \times 8.75} = 88.7\%$. Since we still need to maintain a consistent output, a reasonable upper bound looks close to $\mu = 9.0$. This means that a range between $[86.2\%, 88.7\%]$ utilisation seems to moderate waiting times effectively even in times of high strain.

```{r "Utilisation rate calculations"}
lower <- smu7 / (c * 9.00)
upper <- smu7 / (c * 8.75)
c(lower, upper)
```

```{r "Summary statistics"}
summary(plotlist$"9") # 86.2% utilisation
summary(plotlist$"8.75") #88.7% utilisation
```

Now, drivers will have a maximum wait-time of $6.3$ and $8.5$ minutes respectively for each boundary, with more than 75% having an overall wait-time that is less than $1$ minute.

We can also consider how all boxplots are heavily right-skewed, demonstrating that a majority of commuters will not have a journey impacted by large wait-times regardless of $\mu$. So, while the median wait-time is decreasing, it is at a far slower rate than the extreme change we see in the outliers.

```{r "Outlier Exploration"}
par(mfrow = c(1, 2))
plot(plotlist$"8.5", xlab="Timestep", ylab="Wait Time (min)", main="Wait Times, Mu = 8.5", cex=0.6)
plot(plotlist$"9", xlab="Timestep", ylab="Wait Time (min)", main="Wait Times, Mu = 9.0", cex=0.6)
```

Looking deeper, a majority of of timesteps have a reasonable wait-time – below about 4-minutes. However, nearing the 2000th timestep there is an extreme anomaly with wait times pushing above 10-minutes. This scenario reflects the huge uncertainties in modelling traffic flow – even though we are below capacity, it's still possible for the system to reach breakdown, although it eventually *will* recover.

We can therefore split the sensitivity of a stable system into two distinctions:

-   The $50th$ and $75th$ percentiles (wait-times for the majority of drivers) seem relatively insensitive to changes in $\mu$;

-   The outlier wait-times remain highly sensitive to changes in $\mu$

Once below about $90\%$ utilisation, there is only a minimal improvement to wait-times for the $50th$ and $75th$ percentiles even for a large change in the release rate. This means that it is critical to find a $\mu$ that balances both a consistent and timed release of vehicles without breaking down under the sort of stresses we can see near the 2000th-timestep. A system that would function best here is one that can dynamically adjust its $\mu$ based on a live wait-time distribution.

### Part 2 - Transient Response

Another important regard is to understand how the queuing system responds to a sudden change in the vehicle arrival pattern. For this we need to develop a new function, identical but including:

-   A parameter for vehicle injection (discrete surge) at timestep P

-   A parameter for a model shift (changing to a new arrival rate and size) at timestep Q

-   A maximum queue length before a temporary third lane is opened

```{r "Transient Simulation"}
#| include: false

# FOR THE SIMULATION BELOW
# alambda refers to TOTAL arrival rate /min
# asize refers to dispersion parameter
# rmu refers to PER LANE departure rate /min

# CHANGES
# defaults to 2 lanes
# threshold : the length of queue (count) before a 3rd temp lane is opened
# injection: a vector c(count of vehicles to inject, timestep of injection)
# shift: a vector c(new alambda, new asize, timestep to switch to new model)

onRampTransientSimulation <- function(steps, alambda, asize, rmu, 
                                      injection = c(0, 0), 
                                      shift = c(0, 0, 0), 
                                      threshold = Inf) {
  set.seed(1234)
  
  queue <- c() # storage for vehicles currently in queue
  incoming <- numeric(steps) # track count of vehicles arrived
  outgoing <- numeric(steps) # track count of vehicles released
  
  wlength <- numeric(steps) # track length of queue in each timestep
  wtimesavg <- numeric(steps) # track AVERAGE waiting times of vehicles released in a timestep
  wtimesall <- c() # track INDIVIDUAL waiting times of vehicles released in a timestep

  for (i in 1:steps) {
    # control temporary lane
    c <- 2 # defaults to 2 lanes if below threshold
    if ((i != 1) && (wlength[i-1] > threshold)) {
      c <- 3
    }
    
    # control model shift
    if(i == shift[3]) {
      alambda <- shift[1]
      asize <- shift[2]
    }
    
    # calculate arrivals
    arate <- rnbinom(1, size=asize, mu=alambda)
    # control vehicle injection
    if(i == injection[2]) {
      arate <- arate + injection[1]
    }
    queue <- c(queue, rep(i, arate))
    incoming[i] <- arate

    # calculate departures
    rrate <- sum(rpois(c, lambda=rmu))
    released <- min(length(queue), rrate)
    
    outgoing[i] <- released
    
    # release cars and track wait times
    if (released > 0) {
      rtimes <- i - queue[1:released]
      wtimesavg[i] <- mean(rtimes)
      wtimesall <- c(wtimesall, rtimes)
      queue <- queue[-(1:released)]
    } else {
      wtimesavg[i] <- 0 
    }
    
    wlength[i] <- length(queue)
  }
  
  return(list(
    incoming = incoming,
    outgoing = outgoing,
    wtimesavg = wtimesavg,
    wtimesall = wtimesall,
    wlength = wlength)
    )
}
```

For this new simulation, we will switch focus to the mean model (8am) and stress-test ontop of it's parameters.

::: caption
Vehicle injection comparison for 300 vehicles at timestep 100
:::

```{r "Testing vechile injetion"}
#| echo: false

sim1 <- onRampTransientSimulation(500, smu7, ssize7, 8.25,
                          injection = c(300, 100), # 300 vehicles, timestep 100
)
sim2 <- onRampTransientSimulation(500, smu7, ssize7, 8.75,
                          injection = c(300, 100), # 300 vehicles, timestep 100
)

plot(cumsum(sim1$incoming), type="l", lty=1, col="red", xlab="Timestep", ylab="Cumulative Count", main="Arrivals and Departures")
lines(cumsum(sim1$outgoing), type="l", col="orange")
lines(cumsum(sim2$outgoing), type="l", col="darkgreen")

legend("bottomright", legend=c("Arrivals", "Departures (8.25)", "Departures (8.75)"), col=c("red", "orange", "darkgreen"), lty=1, bty="n")
abline(v=250, col="darkgreen", lty=3)
abline(v=450, col="orange", lty=3)
```

For a system in stability, the above arrival and departure lines need to converge at all times. If we induce a sudden step change change, we can see how long it takes for this equilibrium to reoccur as a way of determining system stress. Comparing the everyday ($\mu = 8.25$, $\rho=94.1\%$), and conservative ($\mu = 8.75$, $\rho=88.7\%$) release rates from earlier, we can use the above plot to see how each model reacts to an injection of $300$ vehicles at $T=100$.

The conservative model quickly recovers, with its departure slope increasing until it meets the arrivals near $T=250$. Since the system is comfortably under capacity, if more vehicles are present it has the ability to increase its output to match. On the other hand, the everyday model converges closer $T=450$, making it $\frac{150}{350}\approx40\%$ slower.

Of course, $300$ vehicles is a somewhat unrealistic number. We can simulate a realistic, but still particularly heavy phase from the preceding intersection by considering a new model:

-   Arrival rate and size corresponding to 8am

-   Running at $85\%$ utilisation

-   A 30-car slug at $T=5$

```{r "Light-phase slug"}
sim <- onRampTransientSimulation(30, smu8, ssize8, rmu8/0.85,
                          injection = c(30, 5))
```

```{r "Plotting transient"}
#| echo: false

par(mfrow = c(1, 2))
plot(cumsum(sim$incoming), type="l", lty=1, col="red", xlab="Time (min)", ylab="Cumulative Count", main="Arrivals and Departures")
lines(cumsum(sim$outgoing), type="l", col="blue")
legend("bottomright", legend=c("Arrivals", "Departures"), col=c("red", "blue"), lty=1, bty="n")
abline(v=5, col="red", lty=3)
abline(v=17, col="blue", lty=3)

plot(sim$wlength, type="l", main="Queue Length", xlab="Time (min)", ylab="Count of Vehicles", col="#b5b5b5")
points(sim$incoming, cex = 0.6, col="#b5b5b5")
legend("topright", legend=c("Queue Length", "Arriving Vehicles"), col="#b5b5b5", lty=c(1, 0), pch=c(NA, 1), bty="n", cex=0.6)
abline(v=5, col="red", lty=3)
abline(v=17, col="blue", lty=3)
```

It takes about 12 minutes for the on-ramp to clear the queue and catch up with its regular input. Remember, this is simulating a heavy phase *ontop* of its typical throughput, so it's still an extreme scenario. But, it represents a quantifiable and realistic way that the system would react to a sudden inflow of vehicles.

```{r "Summary Statistics"}
summary(sim$wtimesavg)
```

Road users are only mildly impacted with $75\%$ of drivers waiting under two minutes, and the maximum wait remaining under three.

### Part 3 - Overall System Impact

Main Roads has a proposed 7-minute improvement in Southbound travel times following the implementation of Smart Freeways. This is calculated from Hester Avenue (Clarkson) to Vincent Street (Leederville), but there are no defined figures from Reid Highway specifically.

We can estimate this through the proportion of lengths:

-   Hester Avenue to Reid Highway is approx. 22km

-   Reid Highway to Vincent Street is approx. 10km

If we assume that time improvements are uniform across the freeway, we can equate a conservative saving of 2 minutes on the Southern segment. For the previous models, we can estimate the net improvement/detriment to commuters by comparing the distribution of on-ramp wait-times with this value.

Since this we are considering the system as a whole here, let's also expand to moderate spillback. Spillback occurs when a signal provides insufficient capacity, forming a queue that affects and blocks throughput of neighbouring intersections. We first need to find the maximum capacity of the road segment:

According to Google Earth, the length of road preceding the on-ramp signal is:

-   Single lane: 330m, and

-   Double lane: 120m

We can also source an effective 8.5m length for a queued passenger vehicle (Main Roads, 2018), accounting for both the length of the vehicle and a gap infront. This leads to a total capacity of $\frac{330+120 \times 2}{8.5}\approx67$ vehicles before spillback occurs.

Consider the maximum, worst-case model (i.e. at 7am) with both previously calculated parameters:

1.  the conservative $88.7\%$ utilisation parameter

2.  a everyday $94.1\%$ utilisation parameter

And a third condition: opening a temporary, third lane when reaching 70% capacity of the network.

```{r}
threshold <- 67 * 0.7 # open temp lane at 70% capacity

par(mfrow = c(1, 2))
# testing 88% utilisation
sim88 <- onRampTransientSimulation(10000, smu7, ssize7, rmu7 / 0.887,
                                   threshold = threshold)
hist(sim88$wtimesavg, main = "7am, p=88.7%", xlab="Waiting Time (minutes)", breaks = 20)
abline(v = 2, lwd = 3, col="red")

# testing 94% utilisation
sim94 <- onRampTransientSimulation(10000, smu7, ssize7, rmu7 / 0.941,
                                   threshold = threshold)
hist(sim94$wtimesavg, main = "7am, p=94.1%", xlab="Waiting Time (minutes)", breaks = 20)
abline(v = 2, lwd = 3, col="red")

# confirming system has not 'spilled' back
max(sim88$wlength)
max(sim94$wlength)
```

The distribution of wait times for both utilisation rates appears exponential. As we near the critical-point (as $\rho \rightarrow 100\%$), the right tail grows fatter and a greater proportion of patrons are waiting more than 2-minutes at the on-ramp. To quantitatively define this, consider a test for the proportion of vehicles with a 2-minute or longer wait at the on-ramp, whose net-travel time across the Southbound Freeway has increased:

```{r "7am Proportion Tests"}
#| include: false

# using 88.7% utilisation rate
prop.test(x = sum(sim88$wtimesavg > 2), 
          n = length(sim88$wtimesavg), 
          p = 0.1, 
          alternative = "greater", 
          conf.level = 0.95)

# using 94.1% utilisation rate
prop.test(x = sum(sim94$wtimesavg > 2), 
          n = length(sim94$wtimesavg), 
          p = 0.1, 
          alternative = "greater", 
          conf.level = 0.95)
```

$$
H_0 = p \leq p_o\
$$

$$
H_a = p > p_o
$$

For a threshold of 10% of drivers with a net increase, (i.e. $p_o=0.1$) and at 95% confidence, see:

```{r "Proportion Test"}
#| echo: false

data.frame(
  P_Value = c("1.00", "0.9999"),
  Proportion = c("0.0218", "0.089"),
  row.names = c("88.7% Utilisation", "94.1% Utilisation")
)
```

For both utilisation rates, we yield an extremely high p-value – we cannot reject the null and so the proportion of drivers seeing an overall increase in journey time is insignificant. For the everyday utilisation rate, we can be $95\%$ confident that the proportion of drivers with a net increase in journey time is at least $8.5\%$. This is, of course below the $10\%$ threshold.

Therefore, as predicted we can maintain a higher utilisation rate as long as the system incorporates feedback to prevent queue breakdown. Due to the inherent variation of traffic flow, it is not reasonable for every driver to individually see a net improvement in their travel time from this on-ramp. For some, and most, road users it is – but not all.

```{r "Summary Statistics"}
summary(sim94$wtimesavg)
mean(sim94$wtimesavg < 2)
```

This does not invalidate drivers who have experienced unreasonable journey times on the Smart Freeways system. We have seen that this is certainly possible. However, when incorporating feedback in the simulation, we can see that the maximum wait shrinks to 4-minutes, a more fair number. Even in this situation, $87.2\%$ (a majority) of drivers will still see a net improvement in their travel time when heading towards the city.

## Conclusion (15 marks)

-   Discussion of results

-   Strengths and limitations

-   Possible extensions (future work)

-   Practical implications

strength - transient simulation

limitation - spillback does not influence arrival rates (drivers may choose other paths)

limitation - assuming travel time improvement from main roads is uniformly distributed (probably more time saved when you are further away than closer)

limitation - resolution of simulation is 1+ minute

limitation - does not consider the impact of more/less traffic entering the freeway by chagning the release rate, which would imapct journey times overall for all drivers

## Contributions

## Appendix

-   Additional plots

-   Extended code

-   Supplementary results

```{r}
# PLOTTING BANK
data <- onRampSimulation(steps=5000, alambda=smu6, asize=ssize6, rmu=(rmu6 - 0), c=2)
incoming <- data$incoming
outgoing <- data$outgoing
wtimesavg <- data$wtimesavg
wtimesall <- data$wtimesall
wlength <- data$wlength

# hysteresis plot
plot(wlength, wtimesavg)

# one more
plot(cumsum(incoming) / cumsum(outgoing))
plot(incoming / outgoing)

# level of service
hist(wtimesall)

plot(cumsum(incoming) / cumsum(outgoing))

# SENSITIVITY THING I DIDNT LIKE
steps <- 1000
rmus <- c(6, 7, 8)
count <- length(rmus)
c <- 2

sims <- runMultiSims(steps, rep(smu7, count), rep(ssize7, count), rmus, c)$sims

par(mfrow = c(1, count))
for (i in seq_along(rmus)) {
  plot(
    sims[[i]]$wlength,
    type = "l",
    xlab = "Timestep",
    ylab = "Count of Vehicles",
    cex.axis = 0.8,
    col = "#b5b5b5",
    main = paste("Mu =", round(rmus[i], 2))
  )
}
```

## References

*2025 Annual Report \| Main Roads Western Australia*. (2025). Wa.gov.au; Main Roads Western Australia.

<https://www.mainroads.wa.gov.au/490d5d/globalassets/about-us/our-organisation/main-roads-annual-report-2025-interactive-version.pdf>

Burmas, G. (2025, February 3). Perth drivers still not convinced by benefits of \$200 million smart freeway. Abc.net.au; ABC News.

<https://www.abc.net.au/news/2025-02-04/perth-smart-freeway-mitchell-kwinana-freeway/104872328>

Righini, G. (2022). *Introduction Probability distributions Birth-and-death process Results for some queuing systems Queuing systems design Queuing theory*.

<https://homes.di.unimi.it/righini/Didattica/Logistica/MaterialeLog/S4%20-%20Queuing%20theory.pdf>

‌Levinson, D. (2021, February 28). Engineering LibreTexts.

<https://eng.libretexts.org/Bookshelves/Civil_Engineering/Fundamentals_of_Transportation/06%3A_Traffice_Control/6.04%3A_Metering_and_Analogs?utm_source=chatgpt.com>

Siegrist, K. (2022, Apr 24). Statistics LibreTexts. [https://stats.libretexts.org/Bookshelves/Probability_Theory/Probability_Mathematical_Statistics_and_Stochastic_Processes\_(Siegrist)/05%3A_Special_Distributions/5.04%3A_Infinitely_Divisible_Distributions](#0)

‌*Traffic Control -Avoiding End of Queue Collisions on High Speed Roads*. (2018). Main Roads.

<https://www.mainroads.wa.gov.au/494be1/globalassets/technical-commercial/working-on-roads/traffic-management/factsheet-avoiding-end-of-queue-collisions.pdf>
