---
title: "A study of on-ramp metering in Perth"
format:
  html:
    css: styles.css
editor: visual
---

| Name              | Student ID |
|-------------------|------------|
| Christian Fazzari | 22220880   |
| Annas Atif        | 22224125   |
| Madhan Jeelaga    | 22815626   |

: *Project Authors*

## Introduction and Background (10 marks)

-   Problem context

-   Objectives

-   Significance

-   Brief literature review (3-5 relevant sources)

Perth's freeway network has undergone major changes in recent years. One of the more controversial upgrades has been the Smart Freeways project; a modification which aims to improve the travel time of commuters from Perth's outer suburbs heading towards the city. This system has been implemented on both the Kwinana Freeway Northbound and the Mitchell Freeway Southbound, controlling the flow of traffic predominantly through digital speed-limit gantries and on-ramp metering. On the Mitchell South, this system saves 7-minutes when travelling from Hester Avenue to Vincent Street in the morning peak (Main Roads, 2025). However, residents closer to the city are generally more impacted by ramp-metering, and, have noticed their morning commute more than doubling in length (ABC, 2025).

In order to weigh the validitity of these claims, this study will analyse the time spent queuing the enter the freeway and compare this with previous research of the time saved on the mainline. We will perform parameter optimisation to adjust and maximise release timing while maintaining stability of the system.

A queuing system with $c$ lanes is stable (i.e. has a controllable load) if its utilisation rate $\rho = \frac{\lambda}{c \times \mu}$ is less than $1$, where $\lambda$ and $\mu$ refer to the mean arrival and departure rates respectively (Righini, 2020). In terms of ramp-metering, this utilisation rate is typically maintained in a range that is safely below capacity as a means of risk aversion, or avoiding a total breakdown of the queue system (Levinson, 2021).

**The primary aim of this study is the determine whether the reported 7-minute travel time savings on the Kwinana Freeway Northbound and Mitchell Freeway Southbound are negated by the increased on-ramp queuing times for local residents. This will be evaluated by developing a queuing simulation based on historical traffic volume data.**

<https://pages.cs.wisc.edu/~dsmyers/cs547/simulation_handout.pdf>

\*\* this is unfinished but just trying to get ideas

```{r "Importing and filtering"}
#| include: false

# Keep these commented as the full files were too large to upload to Github
# Importing datasets
# raw23 <- read.csv("2023-03.csv")
# raw25 <- read.csv("2025-03.csv")

# Filtering for on-ramp segment
# ramp25 <- raw25[raw25$MLinkID == 5725, ]
# ramp23 <- raw23[raw23$MLinkID == 5725, ]

# Filtering for the freeway segment directly following the on-ramp
# fwy25 <- raw25[raw25$MLinkID == 183, ]
# fwy23 <- raw23[raw23$MLinkID == 183, ]

# Simplified import commands
ramp25 <- read.csv("ramp25.csv")
ramp23 <- read.csv("ramp23.csv")

fwy25 <- read.csv("fwy25.csv")
fwy23 <- read.csv("fwy23.csv")
```

```{r "Date and time modifications"}
#| include: false

library(lubridate)

# Converting datetime strings to date objects
ramp23$Date.Time.AWST <- ymd_hm(ramp23$Date.Time.AWST)

# Adding day of the week column (1=sunday, 7=saturday)
ramp23$Weekday <- wday(ramp23$Date.Time.AWST)

# Adding hour of the day column
ramp23$Hour <- hour(ramp23$Date.Time.AWST)

# Adding 15-min segment identifier of column
ramp23$Period <- format(ramp23$Date.Time.AWST, "%H:%M")

# Removing weekends
ramp23.wk <- ramp23[ramp23$Weekday %in% c(2, 3, 4, 5, 6), ]

# Filtering for only the morning peak (6am-10am)
ramp23.pk.wk <- ramp23.wk[6 <= ramp23.wk$Hour & ramp23.wk$Hour < 10, ]

# Removing public holidays (March 6 is the onyl date - Labour Day)
ramp23.pk.wk <- ramp23.pk.wk[date(ramp23.pk.wk$Date.Time.AWST) != "2023-03-06", ]
```

## Methodology (25 marks)

-   Assumptions and model formulation

-   Description of data (if applicable)

-   Justification of modelling choices

Main Roads have a large online repository of traffic volume measurements from 2013 to the present day. Each entry is a 15-minute interval which can be accessed for any road segment using its unique MLinkID. We can filter for the on-ramp of interest using its ID 5725:

```{r "Filtered datset"}
head(ramp23[, c(-6, -7, -8, -9, -10)])
```

> Here, 'Volume' refers to the count of vehicles, and 'TravelTimeMinutes' the average time to travel that particular road segment.

TALK ABOUT MEMORYLESS PROPERTY AS A FALLBACK/ASSUMPTION

In order to construct a queuing model, we can use this dataset to simulate the incoming volume of traffic during the morning peak. Typically, this model would be formed on the Poisson distribution because of its memoryless property and ease of implementation.

> It is important to discuss the memoryless property of the poisson process as this implies that the probability of future events is completely **independent** of the past as the process is unable to 'mathematically determine' the time that has passed since the last event. This makes it an ideal process to use as we can model extreme conditions such as traffic accidents whilst ensuring that they remain completely independent of the other events.

However, a Poisson distribution assumes an equal mean and variance, so it's important that verify this assumption. To visualise, let's filter the above for weekdays, the peak hours, and without public holidays (eg 06/03/23 : Labour Day), yielding the following:

```{r "Aggregating average daily volumes"}
#| include: false

smeans <- aggregate(Volume ~ Period, 
                           data = ramp23.pk.wk, 
                           FUN = mean)

# getting max volume
smax <- max(smeans$Volume)
maxindex <- which(smeans$Volume == smax)

# getting mean volume
smean <- mean(smeans$Volume)

# getting min volume
smin <- min(smeans$Volume)
minindex <- which(smeans$Volume == smin)
```

```{r "Volume barplot"}
#| echo: false

bp <- barplot(smeans$Volume, names.arg = smeans$Period, col="#abbf9d", xlab = "Time of Morning", ylab = "Mean Volume of Period", main="Volume of Merging Traffic, March 2023")

abline(v = bp[1], col = "#252e1f", lwd = 3, lty = 2)
text(x = bp[1], y = smax, labels = round(smeans$Volume[1], 2), pos = 3, xpd = TRUE)

abline(v = bp[5], col = "#252e1f", lwd = 3, lty = 2)
text(x = bp[5], y = smax, labels = round(smeans$Volume[5], 2), pos = 3, xpd = TRUE)

abline(v = bp[9], col = "#252e1f", lwd = 3, lty = 2)
text(x = bp[9], y = smax, labels = round(smeans$Volume[9], 2), pos = 3, xpd = TRUE)
```

The volume of traffic entering the freeway varies heavily over the course of the morning. It would be unreasonable to fit a single model to this level of variation, so instead consider a piecewise approach. Models that account for the minimum, maximum and mean volume each morning will be constructed, correlating to the times 6am, 7am and approximately 8am.

We'll focus on the 6:00-6:15am interval for now:

```{r "Parameters for 6am model"}
#| echo: false
smu6  <- mean(ramp23.pk.wk[ramp23.pk.wk$Period == "06:00", 4])
svar6 <- var(ramp23.pk.wk[ramp23.pk.wk$Period == "06:00", 4])

data.frame (
  Mean = smu6,
  Variance = svar6,
  row.names = c("6am (min)")
)
```

This is an overdispersed dataset where the variance of volume is considerably higher than its mean. This is true throughout the entire dataset, meaning a Poisson distribution would be unsuitable as we do not satisfy the assumption of equidispersion. In order to provide a better fit, we can consider a Negative Binomial where both a mean and a size parameter (for dispersion) can be selected. (NEED TO TALK ABOUT HOW MEMORYLESS IS STILL INVALID AND ALSO ASSUMPTION OF INDEPENDENCE REMAINS)

> It is important to note that, whilst we eliminate the memoryless aspect of the Poisson Process, we still retain independence with the Negative Binomial distribution. In fact, it makes a better proposition compared to the Poisson Process, as we are able to handle overdispersion in the data set which allows for more accurate modelling of accidents, and in high congestion traffic zones which makes it ideal for this studies case.

We can calculate this size parameter $r$ as follows:

$$r = \frac{\mu^2}{\sigma^2 - \mu}$$

```{r "Calculating 6am disperion factor"}
#| include: false
ssize6 <- (smu6^2) / (svar6 - smu6)
```

To see the improvement in fit, let's simulate a Poisson and a Negative Binomial with these parameters and compare with the original dataset.

```{r "Simulating goodness of fit improvement"}
sim1 <- rnbinom(10000, size = ssize6, mu = smu6)
sim2 <- rpois(10000, lambda = smu6)
data.frame (
  Mean = c(smu6, mean(sim1), mean(sim2)),
  Variance = c(svar6, var(sim1), mean(sim2)),
  row.names = c("Dataset", "NBinom Sim", "Poisson Sim  ")
)
```

The Negative Binomial is a much better fit to the original dataset. However, a new concern arises: for the purpose of simulation, we wish to simulate vehicle-counts on a per-minute basis. This will help to increase the granularity of our queuing time results. One possible solution (*Method 1*) is to simply divide the volume data by 15mins and fit a new model. If we consider this case, our mean will scale by a factor of $\frac{1}{15}$, and the variance by a factor of $\frac{1}{15^2}$:

```{r "Method 1 of reducing to Vehicles/Min"}
#| echo: false
data.frame(
  Mean = smu6 / 15,
  Variance = svar6 / 15^2,
  Size = ((smu6/15)^2) / ((svar6/(15^2)) - smu6/15),
  row.names = c("Vehicles/min")
  )
```

Since this is now underdispersed, a Negative Binomial will no longer fit (hence the negative size). It remains unbiased, but erases so much of the original variance that it is completely irrelevant. An alternative solution (*Method 2*) that maintains our original variation is to use the divisible property of the Negative Binomial. We first form the model, then scale its parameters to maintain a dispersion that remains in proportion to the new, lower mean.

> This occurs because the Negative Binomial is formed as a Compound Poisson (Siegrist, 2022), so if $X=X_1 + X_2 +...+X_n \sim NBinom(k, p)$ then we can model each $X_i$ as $NBinom(\frac{k}{n}, p)$ with the same probability $p$. Extrapolating this $(_{successes}, \space _{probability})$ form to the $(_{mean}, \space _{size})$ form yields the parameters $(\frac{\mu}{n}, \space \frac{r}{n})$.

```{r Compare methods 1 and 2 of reducing to Vehicles/Min}
#| echo: false

data.frame(
  Mean = c(smu6, smu6/15, smu6/15),
  Size = c(ssize6, ((smu6/15)^2) / ((svar6/(15^2)) - smu6/15), ssize6/15),
  row.names = c("Vehicles/15min", "Vehicles/min (Method 1)", "Vehicles/min (Method 2)")
)
```

We can extend this idea from the 6am model to all three models.

------------------------------------------------------------------------

::: caption
Disaggregated Model Parameters (vehicles per min)
:::

```{r "Updating 6am and calculating 7am and 8am parameters to Vehicles/Min"}
#| include: false

# 6am models
# leave as the same

# 7am models
smu7  <- mean(ramp23.pk.wk[ramp23.pk.wk$Period == "07:00", 4])
svar7 <- var(ramp23.pk.wk[ramp23.pk.wk$Period == "07:00", 4])
ssize7 <- ((smu7^2) / (svar7 - smu7))

# 8am models
smu8  <- mean(ramp23.pk.wk[ramp23.pk.wk$Period == "08:00", 4])
svar8 <- var(ramp23.pk.wk[ramp23.pk.wk$Period == "08:00", 4])
ssize8 <- ((smu8^2) / (svar8 - smu8))

# update to vehicles / 15 minute
# we no longer use variance! DO NOT TOUCHHHH VARIANCE
smu6 <- smu6 / 15
ssize6 <- ssize6 / 15

smu7 <- smu7 / 15
ssize7 <- ssize7 / 15

smu8 <- smu8 / 15
ssize8 <- ssize8 / 15
```

```{r "Showing parameters for all models"}
#| echo: false

data.frame(
  Mean = c(smu6, smu7, smu8),
  Size = c(ssize6, ssize7, ssize8),
  row.names = c("6am (min)", "7am (max)", "8am (mean)")
)
```

## Model Implementation and Results (45 marks)

Primary points:

-   Implementation (code integrated with explanation)

-   Statistical summaries and visualisations

-   Interpretation in context

Secondary points:

-   Extend to multiple servers (M/M/c) and compare

-   Non-Exponential Service Times (M/G/1)

-   Compare simulation results with analytical formulas (where available)

-   Investigate waiting time distributions

-   Time-varying arrivals (peak vs off-peak) using piecewise arrival rates

-   Conduct sensitivity analysis on arrival and service rates

```{r "Injection"}
#| include: false

yes <- 2

# if(i == 200) {
#   arrivals <- arrivals + 100 # TESTING VEHICLE INJECTION
#   # TESTING PIECEWISE
#   # lambda <- smu7
#   # dispersion <- ssize7
# }

# add another thing to add a temp lane when queue is a certain length
# maybe add optional parameters at end? if you dont input them then sim ignores it 

# FOR THE SIMULATION BELOW
# alambda refers to TOTAL arrival rate /min
# asize refers to dispersion parameter
# rmu refers to PER LANE departure rate /min
# c refers to number of lanes
```

For the simulation, we can develop a queuing model based on the parameters described above. For each model, consider the critical point where the utilisation parameter $\rho$ is exactly one. Rearranging in terms of the release rate, we find $\mu = \frac{1 \times c}{\lambda}$ whichs allows us to determine the largest timing possible for each system to remain stable.

As 2 lanes are present at the Reid Highway on-ramp, we set $c=2$ and form an M/D/2 queue.

```{r "Calculating departure rates"}
#| echo: false

c <- 2

rmu6 <- smu6/c
rmu7 <- smu7/c
rmu8 <- smu8/c

data.frame(
  Best = c(smu6, rmu6),
  Worst = c(smu7, rmu7),
  Average = c(smu8, rmu8),
  row.names = c("Arrival Rate", "Departure Rate")
)
```

Below is the function we have designed for the simulation:

```{r "Primary Simulation"}
onRampSimulation <- function(steps, alambda, asize, rmu, c) {
  set.seed(1234)
  
  queue <- c() # storage for vehicles currently in queue
  incoming <- numeric(steps) # track count of vehicles arrived
  outgoing <- numeric(steps) # track count of vehicles released
  
  wlength <- numeric(steps) # track length of queue in each timestep
  wtimesavg <- numeric(steps) # track AVERAGE waiting times of vehicles released in a timestep
  wtimesall <- c() # track INDIVIDUAL waiting times of vehicles released in a timestep

  for (i in 1:steps) {
    # calculate arrivals
    arate <- rnbinom(1, size=asize, mu=alambda)
    
    queue <- c(queue, rep(i, arate))
    incoming[i] <- arate
    
    # calculate departures
    rrate <- sum(rpois(c, lambda=rmu))
    released <- min(length(queue), rrate)
    
    outgoing[i] <- released
    
    # release cars and track wait times
    if (released > 0) {
      rtimes <- i - queue[1:released]
      wtimesavg[i] <- mean(rtimes)
      wtimesall <- c(wtimesall, rtimes)
      queue <- queue[-(1:released)]
    } else {
      wtimesavg[i] <- 0 
    }
    
    wlength[i] <- length(queue)
  }
  
  return(list(
    incoming = incoming,
    outgoing = outgoing,
    wtimesavg = wtimesavg,
    wtimesall = wtimesall,
    wlength = wlength)
    )
}
```

```{r "Multi-Sim Function"}
#| include: false

# easy method to run mutliple simulations
runMultiSims <- function(steps, alambdas, asizes, rmus, c){
  if (length(rmus) != length(alambdas) || length(alambdas) != length(asizes))
    stop("Must have an equal count for each parameter.")
  
  sims <- list()

  for (i in seq_along(rmus)) {
    sims[[i]] <- onRampSimulation(
      steps = steps, 
      alambda = alambdas[i],
      asize = asizes[i],
      rmu = rmus[i], 
      c = c
    )
  }
  
  maxlength <- 0
  for (sim in sims) {
    maxlength <- max(sim$wlength, maxlength)
  }
  
  data <- list(
    sims = sims,
    maxlength = maxlength
  )
  
  return(data)
}
```

### Investigating utilisation rate

We can run the above simulation for the minimum (best, 6am), maximum (worst, 7am) and mean (average, 8am) models:

::: caption
Best, Worst and Average Case Simulations (at critical points)
:::

```{r "Visualising Stability"}
#| echo: false
alambdas <- c(smu6, smu7, smu8)
asizes <- c(ssize6, ssize7, ssize8)
rmus <- c(rmu6, rmu7, rmu8)

sims <- runMultiSims(10000, alambdas, asizes, rmus, 2)
maxqueue <- sims$maxlength
sims <- sims$sims # can just rassign

par(mfrow=c(1, length(rmus)), mar=c(4, 4, 2, 1))

for (i in seq_along(rmus)) {
  plot(
    sims[[i]]$wlength,
    type = "l",
    xlab = "Timestep",
    ylab = "Count of Vehicles",
    ylim = c(0, maxqueue),
    cex.axis = 0.8,
    col = "#b5b5b5",
    main = paste("Lambda =", round(alambdas[i], 2), ", Mu =", round(rmus[i], 2))
  )
}

paste("Maximum length:", maxqueue)

```

While in all cases there is theoretical stability (i.e. the models all return to 0), we can see that a utilisation rate of exactly $1$ is completely unreasonable in a real-world scenario. The 7am model particularly balloons to a queue of just under 1200 vehicles long, with the others reaching more than 400 vehicles each. In reality, a utilisation rate closer to 90% is generally maintained to balance throughput with wait times.

We can use this information to investigate the systems overall impact in travel-times.

------------------------------------------------------------------------

### Investigating net travel-time impact

As mentioned, Main Roads has a proposed 7-minute improvement in Southbound travel times following the implementation of Smart Freeways. This is calculated from Hester Avenue (Clarkson) to Vincent Street (Leederville), but there are no defined figures from Reid Highway specifically.

We can estimate this through the proportion of lengths:

-   Hester Avenue to Reid Highway is approx. 22km

-   Reid Highway to Vincent Street is approx. 10km

If we assume that time improvements are uniform across the freeway, we can equate a conservative saving of 2 minutes on the Southern segment. For the following models, we can estimate the net improvement/detriment to commuters by comparing the distribution of on-ramp wait-times with this value.

Consider the maximum, worst-case model (i.e. at 7am):

```{r "7am Analysis"}
#| echo: false

par(mfrow = c(1, 2))

# testing 90% utilisation
sim90 <- onRampSimulation(1000, smu7, ssize7, rmu7 / 0.9, 2)
hist(sim90$wtimesavg, main = "7am, p=90%", xlab="Waiting Time (minutes)", breaks = 5)
abline(v = 2, lwd = 3, col="red")

# testing 95% utilisation
sim95 <- onRampSimulation(1000, smu7, ssize7, rmu7 / 0.95, 2)
hist(sim95$wtimesavg, main = "7am, p=95%", xlab="Waiting Time (minutes)", breaks = 10)
abline(v = 2, lwd = 3, col="red")
```

The distribution of wait times for both utilisation rates appears exponential. As we near the critical-point (as $\rho \rightarrow 1$), the right tail grows fatter and a greater proportion of patrons are waiting more than 2-minutes at the on-ramp. To quantitatively define this, consider a test for the proportion of vehicles with a 2-minute or longer wait at the on-ramp, whose net-travel time across the Southbound Freeway has increased:

```{r "7am Proportion Tests"}
#| include: false

# using 90% utilisation rate
prop.test(x = sum(sim90$wtimesavg > 2), 
          n = length(sim90$wtimesavg), 
          p = 0.1, 
          alternative = "greater", 
          conf.level = 0.95)

# using 95% utilisation rate
prop.test(x = sum(sim95$wtimesavg > 2), 
          n = length(sim95$wtimesavg), 
          p = 0.1, 
          alternative = "greater", 
          conf.level = 0.95)
```

$$
H_0 = p \leq p_o\
$$

$$
H_a = p > p_o
$$

For a threshold of 10% (i.e. $p_o=0.1$) and at 95% confidence, we yield:

```{r "Proportion Test"}
#| echo: false

data.frame(
  P_Value = c("1.00", "8.714e-16"),
  Proportion = c("0.034", "0.176"),
  row.names = c("90% Utilisation", "95% Utilisation")
)
```

For a utilisation rate of 90%, we yield an extremely high p-value – we cannot reject the null and so the proportion of drivers seeing an overall increase in journey time is insignificant. Alternatively, for a 95% utilisation rate we see a p-value much lower than 0.05. Here, we are 95% confident that the true proportion of drivers waiting more than 2-minutes is at least 15.7%, a significant result.

```{r "95% Utilisation Summary"}
summary(sim95$wtimesavg)
```

A majority of drivers are still waiting under 1 minute

------------------------------------------------------------------------

### Transient response analysis

ADD SPILLBACK TO EVERYTHING ABOVE

**Spillback Investigation**

Length of single lane - 330m

Length of double lane - 120m

Combined 'single lane equivalent' length - 570m

FIND SOURCES FOR INFO

<https://www.mainroads.wa.gov.au/494be1/globalassets/technical-commercial/working-on-roads/traffic-management/factsheet-avoiding-end-of-queue-collisions.pdf>

A passenger vehicle has an effective length of about 8.5m according to MainRoads.

Therefore our capacity before spillback occurs is approximately 570/8.5 = 67 passenger cars.

```{r}
# PLOTTING BANK
data <- onRampSimulation(steps=5000, alambda=smu6, asize=ssize6, rmu=rmu6, c=2)
incoming <- data$incoming
outgoing <- data$outgoing
wtimesavg <- data$wtimesavg
wtimesall <- data$wtimesall
wlength <- data$wlength

par(mfrow = c(1, 2))
# queue length plot
plot(wlength, type="l", xlab="Timestep", ylab="Count of Vehicles", cex.axis = 0.8, col="#b5b5b5", main="Length of Queue")
# abline(v = 300, lwd=2, col="#737373")

# time spent waiting plot
plot(wtimesavg, type="l", xlab="Timestep", ylab="Time Spent (mins)", cex.axis = 0.8, col="#b5b5b5", main = "Timing of Queue")
# abline(v = 300, lwd=2, col="#737373")
lines(cumsum(wtimesavg)/seq_along(wtimesavg), lwd=2)
# lines(movingavg, col="blue", lwd=2)

# hysteresis plot
plot(wlength, wtimesavg)

# incoming vs outgoing
plot(cumsum(incoming), type="l", col="red", xlab="Timestep", ylab="Cumulative Count", main="Arrivals and Departures")
lines(cumsum(outgoing), type="l", col="blue")
legend("bottomright", legend=c("Arrivals", "Departures"), col=c("red", "blue"), lty=1, bty="n")
# t.test(lm(cumsum(outgoing[1:100]) ~ seq(1, 100)), lm(cumsum(outgoing[200:300]) ~ seq(200:300)), paired = TRUE)

# another incoming vs outgoing
plot(cumsum(outgoing), cumsum(incoming), type="l")

# one more
plot(cumsum(incoming) / cumsum(outgoing))
plot(incoming / outgoing)

# level of service
hist(wtimesall)
```

Conclusion (15 marks)

-   Discussion of results

-   Strengths and limitations

-   Possible extensions (future work)

-   Practical implications

## Contributions

## Appendix

-   Additional plots

-   Extended code

-   Supplementary results

## References

*2025 Annual Report \| Main Roads Western Australia*. (2025). Wa.gov.au; Main Roads Western Australia.

<https://www.mainroads.wa.gov.au/490d5d/globalassets/about-us/our-organisation/main-roads-annual-report-2025-interactive-version.pdf>

Burmas, G. (2025, February 3). Perth drivers still not convinced by benefits of \$200 million smart freeway. Abc.net.au; ABC News.

<https://www.abc.net.au/news/2025-02-04/perth-smart-freeway-mitchell-kwinana-freeway/104872328>

Righini, G. (2022). *Introduction Probability distributions Birth-and-death process Results for some queuing systems Queuing systems design Queuing theory*.

<https://homes.di.unimi.it/righini/Didattica/Logistica/MaterialeLog/S4%20-%20Queuing%20theory.pdf>

‌Levinson, D. (2021, February 28). Engineering LibreTexts.

<https://eng.libretexts.org/Bookshelves/Civil_Engineering/Fundamentals_of_Transportation/06%3A_Traffice_Control/6.04%3A_Metering_and_Analogs?utm_source=chatgpt.com>

Siegrist, K. (2022, Apr 24). Statistics LibreTexts. [https://stats.libretexts.org/Bookshelves/Probability_Theory/Probability_Mathematical_Statistics_and_Stochastic_Processes\_(Siegrist)/05%3A_Special_Distributions/5.04%3A_Infinitely_Divisible_Distributions](#0)

‌
