---
title: "A study of on-ramp metering in Perth"
format:
  html:
    css: styles.css
editor: visual
---

| Name              | Student ID |
|-------------------|------------|
| Christian Fazzari | 22220880   |
| Annas Atif        | 22224125   |
| Madhan Jeelaga    | 22815626   |

: *Project Authors*

## Introduction and Background (10 marks)

-   Problem context

-   Objectives

-   Significance

-   Brief literature review (3-5 relevant sources)

Perth's freeway network has undergone major changes in recent years. One of the more controversial upgrades has been the Smart Freeways project; a modification which aims to improve the travel time of commuters from Perth's outer suburbs heading towards the city. This system has been implemented on both the Kwinana Freeway Northbound and the Mitchell Freeway Southbound, controlling the flow of traffic predominantly through digital speed-limit gantries and on-ramp metering. On the Mitchell South, this system saves 7-minutes when travelling from Hester Avenue to Vincent Street in the morning peak (Main Roads, 2025). However, residents closer to the city are generally more impacted by ramp-metering, and, have noticed their morning commute more than doubling in length (ABC, 2025).

In order to weigh the validitity of these claims, this study will analyse the time spent queuing the enter the freeway and compare this with previous research of the time saved on the mainline. We will perform parameter optimisation to adjust and maximise release timing while maintaining stability of the system.

A queuing system with $c$ lanes is stable (i.e. has a controllable load) if its utilisation rate $\rho = \frac{\lambda}{c \times \mu}$ is less than $1$, where $\lambda$ and $\mu$ refer to the mean arrival and departure rates respectively (Righini, 2020). In terms of ramp-metering, this utilisation rate is typically maintained in a range that is safely below capacity as a means of risk aversion, or avoiding a total breakdown of the queue system (Levinson, 2021).

<https://pages.cs.wisc.edu/~dsmyers/cs547/simulation_handout.pdf>

\*\* this is unfinished but just trying to get ideas

```{r "Importing and filtering"}
#| include: false

# Keep these commented as the full files were too large to upload to Github
# Importing datasets
# raw23 <- read.csv("2023-03.csv")
# raw25 <- read.csv("2025-03.csv")

# Filtering for on-ramp segment
# ramp25 <- raw25[raw25$MLinkID == 5725, ]
# ramp23 <- raw23[raw23$MLinkID == 5725, ]

# Filtering for the freeway segment directly following the on-ramp
# fwy25 <- raw25[raw25$MLinkID == 183, ]
# fwy23 <- raw23[raw23$MLinkID == 183, ]

# Simplified import commands
ramp25 <- read.csv("ramp25.csv")
ramp23 <- read.csv("ramp23.csv")

fwy25 <- read.csv("fwy25.csv")
fwy23 <- read.csv("fwy23.csv")
```

```{r "Date and time modifications"}
#| include: false

library(lubridate)

# Converting datetime strings to date objects
ramp23$Date.Time.AWST <- ymd_hm(ramp23$Date.Time.AWST)

# Adding day of the week column (1=sunday, 7=saturday)
ramp23$Weekday <- wday(ramp23$Date.Time.AWST)

# Adding hour of the day column
ramp23$Hour <- hour(ramp23$Date.Time.AWST)

# Adding 15-min segment identifier of column
ramp23$Period <- format(ramp23$Date.Time.AWST, "%H:%M")

# Removing weekends
ramp23.wk <- ramp23[ramp23$Weekday %in% c(2, 3, 4, 5, 6), ]

# Filtering for only the morning peak (6am-10am)
ramp23.pk.wk <- ramp23.wk[6 <= ramp23.wk$Hour & ramp23.wk$Hour < 10, ]

# Removing public holidays (March 6 is the onyl date - Labour Day)
ramp23.pk.wk <- ramp23.pk.wk[date(ramp23.pk.wk$Date.Time.AWST) != "2023-03-06", ]
```

## Methodology (25 marks)

-   Assumptions and model formulation

-   Description of data (if applicable)

-   Justification of modelling choices

Main Roads have a large online repository of traffic volume measurements from 2013 to the present day. Each entry is a 15-minute interval which can be accessed for any road segment using its unique MLinkID. We can filter for the on-ramp of interest using its ID 5725:

```{r "Filtered datset"}
head(ramp23[, c(-6, -7, -8, -9, -10)])
```

> Here, 'Volume' refers to the count of vehicles, and 'TravelTimeMinutes' the average time to travel that particular road segment.

TALK ABOUT MEMORYLESS PROPERTY AS A FALLBACK/ASSUMPTION

In order to construct a queuing model, we can use this dataset to simulate the incoming volume of traffic during the morning peak. Typically, this model would be formed on the Poisson distribution because of its memoryless property and ease of implementation. However, a Poisson distribution assumes an equal mean and variance, so it's important that verify this assumption. To visualise, let's filter the above for weekdays, the peak hours, and without public holidays (eg 06/03/23 – Labour Day), yielding the following:

```{r "Aggregating average daily volumes"}
#| include: false

smeans <- aggregate(Volume ~ Period, 
                           data = ramp23.pk.wk, 
                           FUN = mean)

# getting max volume
smax <- max(smeans$Volume)
maxindex <- which(smeans$Volume == smax)

# getting mean volume
smean <- mean(smeans$Volume)

# getting min volume
smin <- min(smeans$Volume)
minindex <- which(smeans$Volume == smin)
```

```{r "Volume barplot"}
#| echo: false

bp <- barplot(smeans$Volume, names.arg = smeans$Period, col="#abbf9d", xlab = "Time of Morning", ylab = "Mean Volume of Period", main="Volume of Merging Traffic, March 2023")

abline(v = bp[1], col = "#252e1f", lwd = 3, lty = 2)
text(x = bp[1], y = smax, labels = round(smeans$Volume[1], 2), pos = 3, xpd = TRUE)

abline(v = bp[5], col = "#252e1f", lwd = 3, lty = 2)
text(x = bp[5], y = smax, labels = round(smeans$Volume[5], 2), pos = 3, xpd = TRUE)

abline(v = bp[9], col = "#252e1f", lwd = 3, lty = 2)
text(x = bp[9], y = smax, labels = round(smeans$Volume[9], 2), pos = 3, xpd = TRUE)
```

The volume of traffic entering the freeway varies heavily over the course of the morning. It would be unreasonable to fit a single model to this level of variation, so instead consider a piecewise approach. Models that account for the minimum, maximum and mean volume each morning will be constructed, correlating to the times 6am, 7am and approximately 8am.

We'll focus on the 6:00-6:15am interval for now:

```{r "Parameters for 6am model"}
#| echo: false
smu6  <- mean(ramp23.pk.wk[ramp23.pk.wk$Period == "06:00", 4])
svar6 <- var(ramp23.pk.wk[ramp23.pk.wk$Period == "06:00", 4])

data.frame (
  Mean = smu6,
  Variance = svar6,
  row.names = c("6am (min)")
)
```

This is an overdispersed dataset where the variance of volume is considerably higher than its mean. This is true throughout the entire dataset, meaning a Poisson distribution would be unsuitable as we do not satisfy the assumption of equidispersion. In order to provide a better fit, we can consider a Negative Binomial where both a mean and a size parameter (for dispersion) can be selected. (NEED TO TALK ABOUT HOW MEMORYLESS IS STILL INVALID AND ALSO ASSUMPTION OF INDEPENDENCE REMAINS) We can calculate this size parameter $r$ as follows:

$$r = \frac{\mu^2}{\sigma^2 - \mu}$$

```{r "Calculating 6am disperion factor"}
#| include: false
ssize6 <- (smu6^2) / (svar6 - smu6)
```

To see the improvement in fit, let's simulate a Poisson and a Negative Binomial with these parameters and compare with the original dataset.

```{r "Simulating goodness of fit improvement"}
sim1 <- rnbinom(10000, size = ssize6, mu = smu6)
sim2 <- rpois(10000, lambda = smu6)
data.frame (
  Mean = c(smu6, mean(sim1), mean(sim2)),
  Variance = c(svar6, var(sim1), mean(sim2)),
  row.names = c("Dataset", "NBinom Sim", "Poisson Sim  ")
)
```

The Negative Binomial is a much better fit to the original dataset. However, a new concern arises: for the purpose of simulation, we wish to simulate vehicle-counts on a per-minute basis. This will help to increase the granularity of our queuing time results. One possible solution (*Method 1*) is to simply divide the volume data by 15mins and fit a new model. If we consider this case, our mean will scale by a factor of $\frac{1}{15}$, and the variance by a factor of $\frac{1}{15^2}$:

```{r "Method 1 of reducing to Vehicles/Min"}
#| echo: false
data.frame(
  Mean = smu6 / 15,
  Variance = svar6 / 15^2,
  Size = ((smu6/15)^2) / ((svar6/(15^2)) - smu6/15),
  row.names = c("Vehicles/min")
  )
```

Since this is now underdispersed, a Negative Binomial will no longer fit (hence the negative size). It remains unbiased, but erases so much of the original variance that it is completely irrelevant. An alternative solution (*Method 2*) that maintains our original variation is to use the divisible property of the Negative Binomial. We first form the model, then scale its parameters to maintain a dispersion that remains in proportion to the new, lower mean.

> This occurs because the Negative Binomial is formed as a Compound Poisson (Siegrist, 2022), so if $X=X_1 + X_2 +...+X_n \sim NBinom(k, p)$ then we can model each $X_i$ as $NBinom(\frac{k}{n}, p)$ with the same probability $p$. Extrapolating this $(_{successes}, \space _{probability})$ form to the $(_{mean}, \space _{size})$ form yields the parameters $(\frac{\mu}{n}, \space \frac{r}{n})$.

```{r Compare methods 1 and 2 of reducing to Vehicles/Min}
#| echo: false

data.frame(
  Mean = c(smu6, smu6/15, smu6/15),
  Size = c(ssize6, ((smu6/15)^2) / ((svar6/(15^2)) - smu6/15), ssize6/15),
  row.names = c("Vehicles/15min", "Vehicles/min (Method 1)", "Vehicles/min (Method 2)")
)
```

We can extend this idea from the 6am model to all three models.

------------------------------------------------------------------------

::: caption
Disaggregated Model Parameters (vehicles per min)
:::

```{r "Updating 6am and calculating 7am and 8am parameters to Vehicles/Min"}
#| include: false

# 6am models
# leave as the same

# 7am models
smu7  <- mean(ramp23.pk.wk[ramp23.pk.wk$Period == "07:00", 4])
svar7 <- var(ramp23.pk.wk[ramp23.pk.wk$Period == "07:00", 4])
ssize7 <- ((smu7^2) / (svar7 - smu7))

# 8am models
smu8  <- mean(ramp23.pk.wk[ramp23.pk.wk$Period == "08:00", 4])
svar8 <- var(ramp23.pk.wk[ramp23.pk.wk$Period == "08:00", 4])
ssize8 <- ((smu8^2) / (svar8 - smu8))

# update to vehicles / 15 minute
# we no longer use variance! DO NOT TOUCHHHH VARIANCE
smu6 <- smu6 / 15
ssize6 <- ssize6 / 15

smu7 <- smu7 / 15
ssize7 <- ssize7 / 15

smu8 <- smu8 / 15
ssize8 <- ssize8 / 15
```

```{r "Showing parameters for all models"}
#| echo: false

data.frame(
  Mean = c(smu6, smu7, smu8),
  Size = c(ssize6, ssize7, ssize8),
  row.names = c("6am (min)", "7am (max)", "8am (mean)")
)
```

## Model Implementation and Results (45 marks)

-   Implementation (code integrated with explanation)

-   Statistical summaries and visualisations

-   Interpretation in context

```{r "Simulating Testing"}
T <- 1000

# average 12 cars per minute release
release <- 12

qlength <- numeric(T)

for (i in 1:T) {
  qlength[i+1] <- qlength[i] + rnbinom(1, size=ssize8, mu=smu8)
  qlength[i+1] <- max(0, qlength[i+1] - release)
}

par(mfrow=c(1, 2))
plot(qlength)


# average 11 cars per minute
release <- 11

qlength <- numeric(T)

for (i in 1:T) {
  qlength[i+1] <- qlength[i] + rnbinom(1, size=ssize8, mu=smu8)
  qlength[i+1] <- max(0, qlength[i+1] - release)
}

plot(qlength)
```

trying new method

```{r}
set.seed(1234)
T <- 500 # no. of timesteps (each is a minute) #### 250 timesteps helps to see stuff better
c <- 2 # no. of lanes

mu <- 6 # mean no. vehicles released PER LANE in one timestep
lambda <- smu8 # mean no. vehicles arriving TOTAL in one timestep
dispersion <- ssize8 # dispersion parameter for arrival rate

queue <- c()
incoming <- c()
outgoing <- c()

wtimes <- list()
wlength <- numeric(T)

for (i in 1:T) {
  # arrivals
  arrivals <- rnbinom(1, size = dispersion, mu = lambda)
  
  if(i == 200) {
    arrivals <- arrivals + 100 # TESTING VEHICLE INJECTION
    # TESTING PIECEWISE
    # lambda <- smu7
    # dispersion <- ssize7
  }
  
  queue <- c(queue, rep(i, arrivals))
  incoming <- c(incoming, arrivals)

  # release parameters
  rrate <- sum(rpois(c, mu))
  rtimes <- c()

  # finding how many cars should be released
  released <- min(length(queue), rrate)
  outgoing <- c(outgoing, released)

  # releasing cars and tracking how long they waited
  if (released > 0) {
    rtimes <- c(rtimes, i - queue[1:released])
    queue <- queue[-(1:released)]
  }

  # record data
  wtimes[[i]] <- rtimes
  wlength[i] <- length(queue)

}

test <- wtimes

for (i in 1:T) {
  wtimes[[i]] <- mean(wtimes[[i]])
}
wtimes <- unlist(wtimes)
mean(wtimes, na.rm = TRUE)
mean(wlength, na.rm = TRUE)

cleanwtimes <- wtimes
cleanwtimes[is.na(cleanwtimes)] <- 0
k <- 20 # 10 averages
movingavg <- filter(cleanwtimes, rep(1/k, k), sides = 1)
```

```{r}
par(mfrow = c(1, 2))
# queue length plot
plot(wlength, type="l", xlab="Timestep", ylab="Count of Vehicles", cex.axis = 0.8, col="#b5b5b5", main="Length of Queue")
# abline(v = 300, lwd=2, col="#737373")

# time spent waiting plot
plot(wtimes, type="l", xlab="Timestep", ylab="Time Spent (mins)", cex.axis = 0.8, col="#b5b5b5", main = "Timing of Queue")
# abline(v = 300, lwd=2, col="#737373")
lines(cumsum(wtimes)/seq_len(T), lwd=2)
# lines(movingavg, col="blue", lwd=2)

# hysteresis plot
plot(wlength, wtimes)

# incoming vs outgoing
plot(cumsum(incoming), type="l", col="red", xlab="Timestep", ylab="Cumulative Count", main="Arrivals and Departures")
lines(cumsum(outgoing), type="l", col="blue")
legend("bottomright", legend=c("Arrivals", "Departures"), col=c("red", "blue"), lty=1, bty="n")
# t.test(lm(cumsum(outgoing[1:100]) ~ seq(1, 100)), lm(cumsum(outgoing[200:300]) ~ seq(200:300)), paired = TRUE)

# another incoming vs outgoing
plot(cumsum(outgoing), cumsum(incoming), type="l")

# one more
plot(cumsum(incoming) / cumsum(outgoing))
plot(incoming / outgoing)

# level of service
hist(wtimes)
```

old method:

```{r}
# T <- 1000 # no. of timesteps (each is a 15min interval)
# c <- 2 # no. of lanes
# 
# mu <- 6 # mean no. vehicles released PER LANE in one timestep
# lambda <- smu8 # mean no. vehicles arriving TOTAL in one timestep
# dispersion <- ssize8 # dispersion parameter for arrival rate
# 
# q1 <- c()
# q2 <- c()
# 
# wtimes <- list()
# q1c <- numeric(T)
# q2c <- numeric(T)
# 
# avg_wait_at_i <- list()
# 
# for (i in 1:T) {
#   # arrivals
#   arrivals <- rnbinom(1, size = dispersion, mu = lambda)
# 
#   q1a <- floor(arrivals / 2)
#   q2a <- ceiling(arrivals /2)
# 
#   q1 <- c(q1, rep(i, q1a))
#   q2 <- c(q2, rep(i, q2a))
# 
#   # release parameters
#   release <- mu
#   timestep.wtimes <- c()
# 
#   # finding how many cars should be released
#   q1r <- min(length(q1), release)
#   q2r <- min(length(q2), release)
# 
#   # releasing cars and tracking how long they waited
#   if (q1r > 0) {
#     timestep.wtimes <- c(timestep.wtimes, i - q1[1:q1r])
#     q1 <- q1[-(1:q1r)]
#   }
#   if (q2r > 0) {
#     timestep.wtimes <- c(timestep.wtimes, i - q2[1:q2r])
#     q2 <- q2[-(1:q2r)]
#   }
# 
#   # record data
#   wtimes[[i]] <- timestep.wtimes
#   q1c[i] <- length(q1)
#   q2c[i] <- length(q2)
# 
#   avg_wait_at_i[i] <- if(length(timestep.wtimes) > 0) mean(timestep.wtimes)
# 
# }
# all_waits <- unlist(wtimes)
# plot(q1c)
# # plot(1:T, avg_wait_at_i)
```

## Conclusion (15 marks)

-   Discussion of results

-   Strengths and limitations

-   Possible extensions (future work)

-   Practical implications

## Contributions

## Appendix

-   Additional plots

-   Extended code

-   Supplementary results

## References

*2025 Annual Report \| Main Roads Western Australia*. (2025). Wa.gov.au; Main Roads Western Australia.

<https://www.mainroads.wa.gov.au/490d5d/globalassets/about-us/our-organisation/main-roads-annual-report-2025-interactive-version.pdf>

Burmas, G. (2025, February 3). Perth drivers still not convinced by benefits of \$200 million smart freeway. Abc.net.au; ABC News.

<https://www.abc.net.au/news/2025-02-04/perth-smart-freeway-mitchell-kwinana-freeway/104872328>

Righini, G. (2022). *Introduction Probability distributions Birth-and-death process Results for some queuing systems Queuing systems design Queuing theory*.

<https://homes.di.unimi.it/righini/Didattica/Logistica/MaterialeLog/S4%20-%20Queuing%20theory.pdf>

‌Levinson, D. (2021, February 28). Engineering LibreTexts.

<https://eng.libretexts.org/Bookshelves/Civil_Engineering/Fundamentals_of_Transportation/06%3A_Traffice_Control/6.04%3A_Metering_and_Analogs?utm_source=chatgpt.com>

Siegrist, K. (2022, Apr 24). Statistics LibreTexts. [https://stats.libretexts.org/Bookshelves/Probability_Theory/Probability_Mathematical_Statistics_and_Stochastic_Processes\_(Siegrist)/05%3A_Special_Distributions/5.04%3A_Infinitely_Divisible_Distributions](#0)

‌
